---
title: "MLB Hits Prediction"
output: html_notebook
---

Analysis using Multivariate, Ctree, and Cluster to predict how many hits any given MLB player should have.

# MLB Hits Predicition Analysis  {.tabset .tabset-fade .tabset-pills}
## MLB Multivarite Regerssion
### Install and load libraries

```{r}
install.packages("tidyverse")

library(tidyverse)
```

### Load and preview data set

```{r}
batting <- read.csv("batting2019.csv")

structure(batting)
```

### Visualization of Average vs. Hits
```{r}
ggplot(data = batting, mapping = aes(x = AVG, y = H)) +
  geom_point(mapping = aes(color = HR)) +
  geom_smooth(method = "lm") +
  labs(title = "Average vs. Hits" ,
         subtitle = "Source: MLB Batting Stats 2019" ,
         x = "Average", y = "Hits")
```

###Correlation: Pairs Plot
```{r}
hB19 <- select(batting, c("AVG", "H"))
pairs(hB19)
```

### Corrlation: calculation
Calculate correlation value [-1, 1]
```{r}
hBCorrelation19 <- cor(batting$AVG, batting$H)
```


### Multivariate Linear Regression 1
Given 2 independent variables, predict the dependent variable.
Given the batting average and slugging percentage, let's predict the player's amount of hits.
```{r}
model191 <- lm(H ~ AVG + SLG, data = batting)
model191
```

### Prediction
Given an average of .305 and slugging of .500, predict the hits.
```{r}
# Create a data frame with AVG and SLG entries
test191 <- data.frame(AVG = c(.305), SLG = c(.500))
# Predict the position with the model and test data
predictedH191 <- predict(model191, test191)
cat("At .305 Average, the Hits is", predictedH191)
```

### Multivariate Linear Regression 2
Given 2 independent variables, predict the dependent variable.
Given the batting average and on-base percentage, let's predict the player's amount of hits.
```{r}
model192 <- lm(H ~ AVG + OBP, data = batting)
model192
```
### Prediction
Given an average of .305 and on-base of .400, predict the hits.
```{r}
# Create a data frame with AVG and OBP entries
test192 <- data.frame(AVG = c(.305), OBP = c(.400))
# Predict the position with the model and test data
predictedH192 <- predict(model192, test192)
cat("At .305 Average, the Hits is", predictedH192)
```

### Multivariate Linear Regression 3
Given 3 independent variables, predict the dependent variable.
Given the batting average and on-base percentage and slugging percentage, let's predict the player's amount of hits.
```{r}
model193 <- lm(H ~ AVG + OBP + SLG, data = batting)
model193
```
### Prediction
Given an average of .305 and on-base of .400 and slugging of .500, predict the hits.
```{r}
# Create a data frame with AVG, OBP and SLG entries
test193 <- data.frame(AVG = c(.305), OBP = c(.400), SLG = c(.500))
# Predict the position with the model and test data
predictedH193 <- predict(model193, test193)
cat("At .305 AVG, the H is", predictedH193)
```

### Verification
Let's see the average hits at .305 batting average.
```{r}
# Output table with AVG of .305
avg30519 <- subset(batting, AVG == .305)
avg30519
# Calculate the avg and sd of Hits when AVG is .305
avgH19 <- mean(avg30519$H)
sdH19 <- sd(avg30519$H)
```
### Load CaTools
```{r}
install.packages("caTools")
library(caTools)
```


### Mulivariate 1
```{r}
# Set the random seed for repeatability
set.seed(123) 
# Split the data into 3:1 ratio
sample191 = sample.split(batting$AB, SplitRatio = .75)
train191 = subset(batting, sample191 == TRUE)
test191 = subset(batting, sample191 == FALSE)
```

```{r}
predict191 <- lm(H ~ AVG, data = test191)
predict191
```

```{r}
# Predict the position with the model and test data
predictedH191 <- predict(predict191, test191)
predictedH191
# Create data frame with actual and predicted values
battingPrediction191 <- data.frame(AVG = test191$AVG, 
                               ActualH = test191$H,
                               PredictedH = predictedH191)
battingPrediction191
```


### Multivariate 2
```{r}
# Set the random seed for repeatability
set.seed(123) 
# Split the data into 3:1 ratio
sample192 = sample.split(batting$AB, SplitRatio = .75)
train192 = subset(batting, sample192 == TRUE)
test192  = subset(batting, sample192 == FALSE)
```

```{r}
predict192 <- lm(H ~ AVG + OBP, data = train191)
predict192
```

```{r}
# Predict the position with the model and test data
predictedH192 <- predict(predict192, test192)
predictedH192
# Create data frame with actual and predicted values
battingPrediction192 <- data.frame(AVG = test191$AVG, 
                               ActualH = test191$H,
                               PredictedH = predictedH192)
battingPrediction192
```

### Multivariate 3
```{r}
# Set the random seed for repeatability
set.seed(123) 
# Split the data into 3:1 ratio
sample193 = sample.split(batting$AB, SplitRatio = .75)
train193 = subset(batting, sample193 == TRUE)
test193  = subset(batting, sample193 == FALSE)
```

```{r}
predict193 <- lm(H ~ AVG + OBP + SLG, data = train193)
predict193
```

```{r}
# Predict the position with the model and test data
predictedH193 <- predict(predict193, test193)
predictedH193
# Create data frame with actual and predicted values
battingPrediction193 <- data.frame(AVG = test193$AVG, 
                               ActualH = test193$H,
                               PredictedH = predictedH193)
battingPrediction193
```

### Visualization
Hits vs. Average
```{r}
ggplot(data = battingPrediction192, mapping = aes(x = AVG)) +
  geom_point(data = battingPrediction192, 
             mapping = aes(y = ActualH),
             color = "blue") +
  geom_point(data = battingPrediction192, 
             mapping = aes(y = PredictedH), 
             color = "red") +
  labs(title = "Hits vs. Average" ,
        subtitle = "Actual (Blue) vs. Predicted (Red)",
        y = "Hits", x = "Average")
```

### Summary
```{r}
summary(model192)
```



### install and Loading Metrics
```{r}
# Using the Metrics library for MSE
install.packages("Metrics")
library(Metrics)
```
### Multivariate 1 Statistical analysis 
```{r}
# Mean Absolute Error
mae191 <- mae(battingPrediction191$ActualH, battingPrediction191$PredictedH)
cat("Mean Absolute Error:", mae191)
```


```{r}
# Mean Squared Error
mse191 <- mse(battingPrediction191$ActualH, battingPrediction191$PredictedH)
cat("Mean Squared Error:", mse191)
```


```{r}
# Root Mean Squared Error
sqrt(mse191)
rmse191 <- rmse(battingPrediction191$ActualH, battingPrediction191$PredictedH)
cat("Root Mean Squared Error:", rmse191)
```

### Multivariate 2 statistical analysis 
```{r}
# Mean Absolute Error
mae192 <- mae(battingPrediction192$ActualH, battingPrediction192$PredictedH)
cat("Mean Absolute Error:", mae192)
```


```{r}
# Mean Squared Error
mse192 <- mse(battingPrediction192$ActualH, battingPrediction192$PredictedH)
cat("Mean Squared Error:", mse192)
```


```{r}
# Root Mean Squared Error
sqrt(mse192)
rmse192 <- rmse(battingPrediction192$ActualH, battingPrediction192$PredictedH)
cat("Root Mean Squared Error:", rmse192)
```


### Multivariate 3 Statistical analysis 
```{r}
# Mean Absolute Error
mae193 <- mae(battingPrediction193$ActualH, battingPrediction193$PredictedH)
cat("Mean Absolute Error:", mae193)
```


```{r}
# Mean Squared Error
mse193 <- mse(battingPrediction193$ActualH, battingPrediction193$PredictedH)
cat("Mean Squared Error:", mse193)
```


```{r}
# Root Mean Squared Error
sqrt(mse193)
rmse193 <- rmse(battingPrediction193$ActualH, battingPrediction193$PredictedH)
cat("Root Mean Squared Error:", rmse193)
```

## MLB Ctree

```{r}
install.packages("tidyverse")
install.packages("party")
```

```{r}
library(tidyverse)
library(party)
```

```{r}
batting <- read.csv("batting2019.csv")
structure(batting)
```




```{r}
MLBtree <- ctree(H ~ AB + OBP + SLG , data = batting)
MLBtree
```

```{r}
plot(MLBtree)
```


```{r}
# Set the random seed for repeatability
set.seed(123) 

# Split the data into 3:1 ratio
sample = sample.split(batting$AB, SplitRatio = .75)
train = subset(batting, sample == TRUE)
test = subset(batting, sample == FALSE)
```



```{r}
predict <- lm(H ~ OBP + AB + SLG, data = test)
predict
```


```{r}
# Predict the position with the model and test data
predictedH <- predict(predict, test)
predictedH

# Create data frame with actual and predicted values
battingPrediction <- data.frame(AVG = test$AVG, 
                               ActualH = test$H,
                               PredictedH = predictedH)

battingPrediction
```

```{r}
# Mean Absolute Error
mae <- mae(battingPrediction$ActualH, battingPrediction$PredictedH)
cat("Mean Absolute Error:", mae)
```

```{r}
testPlayer <- data.frame(AB = as.integer(c(500)),
                          OBP = as.double(c(.320)),
                          SLG = as.double(c(.400)))
predictedPlayer <- predict(MLBtree, newdata = testPlayer)
predictedPlayer
```


###  aditional  variables 


```{r}
MLBtree1 <- ctree(H ~ AB + OBP + SLG + AVG ,  data = batting)
MLBtree1
```


```{r}
plot(MLBtree1)
```


```{r}
# Set the random seed for repeatability
set.seed(123) 

# Split the data into 3:1 ratio
sample1 = sample.split(batting$AB, SplitRatio = .75)
train1 = subset(batting, sample1 == TRUE)
test1 = subset(batting, sample1 == FALSE)
```


```{r}
predict1 <- lm(H ~ AVG + OBP + AB + SLG, data = test1)
predict1
```


```{r}
# Predict the position with the model and test data
predictedH1 <- predict(predict1, test1)
predictedH1

# Create data frame with actual and predicted values
battingPrediction1 <- data.frame(AVG = test1$AVG, 
                               ActualH = test1$H,
                               PredictedH = predictedH1)

battingPrediction1
```

```{r}
# Mean Absolute Error
mae1 <- mae(battingPrediction1$ActualH, battingPrediction1$PredictedH)
cat("Mean Absolute Error:", mae1)
```

```{r}
testPlayer1 <- data.frame(AB = as.integer(c(500)),
                          OBP = as.double(c(.320)),
                          SLG = as.double(c(.400)),
                          AVG = as.double(c(.300)))
predictedPlayer1 <- predict(MLBtree1, newdata = testPlayer1)
predictedPlayer1
```


### Additional  Variables part 3

```{r}
MLBtree11 <- ctree(H ~ AVG + OBP,  data = batting)
MLBtree11
```


```{r}
plot(MLBtree11)
```


```{r}
# Set the random seed for repeatability
set.seed(123) 

# Split the data into 3:1 ratio
sample11 = sample.split(batting$AB, SplitRatio = .75)
train11 = subset(batting, sample11 == TRUE)
test11 = subset(batting, sample11 == FALSE)
```


```{r}
predict11 <- lm(H ~ AVG + OBP, data = test11)
predict11
```


```{r}
# Predict the position with the model and test data
predictedH11 <- predict(predict11, test11)
predictedH11

# Create data frame with actual and predicted values
battingPrediction11 <- data.frame(AVG = test11$AVG, 
                               ActualH = test11$H,
                               PredictedH = predictedH11)

battingPrediction11
```
```{r}
# Mean Absolute Error
mae11 <- mae(battingPrediction11$ActualH, battingPrediction11$PredictedH)
cat("Mean Absolute Error:", mae11)
```

```{r}
# Mean Squared Error
mse11 <- mse(battingPrediction11$ActualH, battingPrediction11$PredictedH)
cat("Mean Squared Error:", mse11)
```

```{r}
# Root Mean Squared Error
sqrt(mse11)
```

```{r}
rmse11 <- rmse(battingPrediction11$ActualH, battingPrediction11$PredictedH)
cat("Root Mean Squared Error:", rmse11)
```





```{r}
testPlayer11 <- data.frame(OBP = as.double(c(.320)),
                          AVG = as.double(c(.300)))
predictedPlayer11 <- predict(MLBtree11, newdata = testPlayer11)
predictedPlayer11
```

## MLB Cluster

### Install and Load packages
```{r}
install.packages("tidyverse")
library(tidyverse)
```

#Load ggplot2
```{r}
library(ggplot2)
```


```{r}
batting <- read.csv("batting2019.csv")
structure(batting)
```


```{r}
ggplot(data = batting, mapping = aes(x = AVG, y = H)) +
  geom_point(mapping = aes(color = HR)) +
  geom_smooth(method = "lm") +
  labs(title = "Batting Average vs Hits" ,
         subtitle = "batting" ,
         x = "AVG")
```


```{r}
set.seed(20)
mlbHits <- data.frame(H = batting$H, AB =batting$AB)
head(mlbHits)
```


```{r}
mlbCluster <- kmeans(mlbHits, 3, nstart = 25)
mlbCluster
```


```{r}
table(mlbCluster$cluster, batting$H)
```


```{r}
mlbHits$Cluster <- as.factor(mlbCluster$cluster)
centroids <- as.data.frame(mlbCluster$centers)
centroids$Cluster <- as.factor(c(1:3))
ggplot(data = mlbHits, mapping = aes(x = H, y = AB, color = Cluster))+
  geom_point()+
  geom_point(data = centroids,aes(x = H, y = AB, fill = Cluster), size = 5, shape = 13) + labs(title = "At-Bats vs. Hits", x = "Hits", y = "At-Bats")
```


```{r}
wss <- numeric(15)
for (k in 1:15)
  wss[k] = sum(kmeans(mlbHits, k, nstart = 25)$withinss)
```


```{r}
wssResults <- data.frame(k = c(1:15), wss = wss)
wssResults
```


```{r}
ggplot(data = wssResults, mapping = aes(x = k, y = wss))+
  geom_point()+
  geom_line()+
  labs(title = "K-means: MLB Hits", x = "K", y = "WSS")
```
