---
title: "R Notebook"
output: html_notebook
---
#Loading Tidyverse
```{r}
install.packages("tidyverse")
library(tidyverse)

```
#Uploading dataset
```{r}
batting <- read.csv("batting2019.csv")

structure(batting)
```
# Visualize the data
Average vs. Hits
```{r}
ggplot(data = batting, mapping = aes(x = AVG, y = H)) +
  geom_point(mapping = aes(color = HR)) +
  geom_smooth(method = "lm") +
  labs(title = "Average vs. Hits" ,
         subtitle = "Source: MLB Batting Stats 2019" ,
         x = "Average", y = "Hits")
```
# Correlation: Pairs Plot
Pair plot to see correlation between variables
```{r}
hB19 <- select(batting, c("AVG", "H"))
pairs(hB19)
```
# Corrlation: calculation
Calculate correlation value [-1, 1]
```{r}
hBCorrelation19 <- cor(batting$AVG, batting$H)
```
# Multivariate Linear Regression 1
Given 2 independent variables, predict the dependent variable.
Given the batting average and slugging percentage, let's predict the player's amount of hits.
```{r}
model191 <- lm(H ~ AVG + SLG, data = batting)
model191
```
# Prediction
Given an average of .305 and slugging of .500, predict the hits.
```{r}
# Create a data frame with AVG and SLG entries
test191 <- data.frame(AVG = c(.305), SLG = c(.500))

# Predict the position with the model and test data
predictedH191 <- predict(model191, test191)
cat("At .305 Average, the Hits is", predictedH191)
```
# Multivariate Linear Regression 2
Given 2 independent variables, predict the dependent variable.
Given the batting average and on-base percentage, let's predict the player's amount of hits.
```{r}
model192 <- lm(H ~ AVG + OBP, data = batting)
model192
```
# Prediction
Given an average of .305 and on-base of .400, predict the hits.
```{r}
# Create a data frame with AVG and OBP entries
test192 <- data.frame(AVG = c(.305), OBP = c(.400))

# Predict the position with the model and test data
predictedH192 <- predict(model192, test192)
cat("At .305 Average, the Hits is", predictedH192)
```
# Multivariate Linear Regression 3
Given 3 independent variables, predict the dependent variable.
Given the batting average and on-base percentage and slugging percentage, let's predict the player's amount of hits.
```{r}
model193 <- lm(H ~ AVG + OBP + SLG, data = batting)
model193
```
# Prediction
Given an average of .305 and on-base of .400 and slugging of .500, predict the hits.
```{r}
# Create a data frame with AVG, OBP and SLG entries
test193 <- data.frame(AVG = c(.305), OBP = c(.400), SLG = c(.500))

# Predict the position with the model and test data
predictedH193 <- predict(model193, test193)
cat("At .305 AVG, the H is", predictedH193)
```
# Verification
Let's see the average hits at .305 batting average.
```{r}
# Output table with AVG of .305
avg30519 <- subset(batting, AVG == .305)
avg30519

# Calculate the avg and sd of Hits when AVG is .305
avgH19 <- mean(avg30519$H)
sdH19 <- sd(avg30519$H)
```
Loading CaTools
```{r}
install.packages("caTools")
library(caTools)
```
# Mulivariate 1
```{r}
# Set the random seed for repeatability
set.seed(123) 

# Split the data into 3:1 ratio
sample191 = sample.split(batting$AB, SplitRatio = .75)
train191 = subset(batting, sample191 == TRUE)
test191 = subset(batting, sample191 == FALSE)
```

```{r}
predict191 <- lm(H ~ AVG, data = test191)
predict191
```

```{r}
# Predict the position with the model and test data
predictedH191 <- predict(predict191, test191)
predictedH191

# Create data frame with actual and predicted values
battingPrediction191 <- data.frame(AVG = test191$AVG, 
                               ActualH = test191$H,
                               PredictedH = predictedH191)

battingPrediction191
```
# Multivariate 2
```{r}
# Set the random seed for repeatability
set.seed(123) 

# Split the data into 3:1 ratio
sample192 = sample.split(batting$AB, SplitRatio = .75)
train192 = subset(batting, sample192 == TRUE)
test192  = subset(batting, sample192 == FALSE)
```

```{r}
predict192 <- lm(H ~ AVG + OBP, data = train191)
predict192
```

```{r}
# Predict the position with the model and test data
predictedH192 <- predict(predict192, test192)
predictedH192

# Create data frame with actual and predicted values
battingPrediction192 <- data.frame(AVG = test191$AVG, 
                               ActualH = test191$H,
                               PredictedH = predictedH192)

battingPrediction192
```
# Multivariate 3
```{r}
# Set the random seed for repeatability
set.seed(123) 

# Split the data into 3:1 ratio
sample193 = sample.split(batting$AB, SplitRatio = .75)
train193 = subset(batting, sample193 == TRUE)
test193  = subset(batting, sample193 == FALSE)
```

```{r}
predict193 <- lm(H ~ AVG + OBP + SLG, data = train193)
predict193
```

```{r}
# Predict the position with the model and test data
predictedH193 <- predict(predict193, test193)
predictedH193

# Create data frame with actual and predicted values
battingPrediction193 <- data.frame(AVG = test193$AVG, 
                               ActualH = test193$H,
                               PredictedH = predictedH193)

battingPrediction193
```
# Visualization
Hits vs. Average
```{r}
ggplot(data = battingPrediction192, mapping = aes(x = AVG)) +
  geom_point(data = battingPrediction192, 
             mapping = aes(y = ActualH),
             color = "blue") +
  geom_point(data = battingPrediction192, 
             mapping = aes(y = PredictedH), 
             color = "red") +
  labs(title = "Hits vs. Average" ,
        subtitle = "Actual (Blue) vs. Predicted (Red)",
        y = "Hits", x = "Average")
```

```{r}
summary(model192)
```
# Loading Metrics
```{r}
# Using the Metrics library for MSE
install.packages("Metrics")
library(Metrics)
```
# Multivariate 1
```{r}
# Mean Absolute Error
mae191 <- mae(battingPrediction191$ActualH, battingPrediction191$PredictedH)
cat("Mean Absolute Error:", mae191)
```
```{r}
# Mean Squared Error
mse191 <- mse(battingPrediction191$ActualH, battingPrediction191$PredictedH)
cat("Mean Squared Error:", mse191)
```
```{r}
# Root Mean Squared Error
sqrt(mse191)
rmse191 <- rmse(battingPrediction191$ActualH, battingPrediction191$PredictedH)
cat("Root Mean Squared Error:", rmse191)
```
# Multivariate 2
```{r}
# Mean Absolute Error
mae192 <- mae(battingPrediction192$ActualH, battingPrediction192$PredictedH)
cat("Mean Absolute Error:", mae192)
```
```{r}
# Mean Squared Error
mse192 <- mse(battingPrediction192$ActualH, battingPrediction192$PredictedH)
cat("Mean Squared Error:", mse192)
```
```{r}
# Root Mean Squared Error
sqrt(mse192)
rmse192 <- rmse(battingPrediction192$ActualH, battingPrediction192$PredictedH)
cat("Root Mean Squared Error:", rmse192)
```
# Multivariate 3
```{r}
# Mean Absolute Error
mae193 <- mae(battingPrediction193$ActualH, battingPrediction193$PredictedH)
cat("Mean Absolute Error:", mae193)
```
```{r}
# Mean Squared Error
mse193 <- mse(battingPrediction193$ActualH, battingPrediction193$PredictedH)
cat("Mean Squared Error:", mse193)
```
```{r}
# Root Mean Squared Error
sqrt(mse193)
rmse193 <- rmse(battingPrediction193$ActualH, battingPrediction193$PredictedH)
cat("Root Mean Squared Error:", rmse193)
```

